\begin{copyrightpage}
\begin{center}
\textbf{Editorial Board}
\end{center}

\begin{center}
David Bressoud \hfill Carl Pomerance\\
Robert Devaney, Chair \hfill Hung-Hsi Wu
\end{center}

\begin{center}
2000 \emph{Mathematics Subject Classification}. Primary 60-02;\\
Secondary 60J10, 60J15, 65C05.
\end{center}

\begin{center}
Research of Gregory Lawler partially supported\\
by the National Science Foundation
\end{center}

\noindent \textsc{Abstract.} Lectures given to undergraduates at the Institute for Advanced Study / Park City Institute on Probability in 1996. Topics include: random walk (simple and self-avoiding), Brownian motion, card shuffling, Markov chains and Markov chain Monte Carlo, spanning trees, and simulation of random walks and models in finance. This book could be used as a supplement to or for a second course after a calculus based probability course.\\

\hrulefill\\

\noindent \textbf{Library of Congress Cataloging-in-Publication
Data}

\noindent Lawler, Gregory F., 1955--

Lectures on contemporary probability / Gregory F. Lawler, Lester N.
Coyle.

\quad p. cm. --- (Student mathematical library, ISSN 1520-9121 ; v.
2)

Includes bibliographical references.

ISBN 0-8218-2029-X (softcover : alk. paper)

1. Stochastic processes. \quad 2. Probabilities. \quad I. Coyle,
Lester N. (Lester Noel), 1967--. \quad II. Title. \quad III. Series.

\noindent QA274.L384 \quad  1999

\noindent 519.2--dc21 \hfill 99-23838

\hfill CIP

\hrulefill

\textbf{Copying and reprinting.} Individual readers of this
publication, and nonprofit libraries acting for them, are permitted
to make fair use of the material, such as to copy a chapter for use
in teaching or research. Permission is granted to quote brief
passages from this publication in reviews, provided the customary
acknowledgment of the source is given.

Republication, systematic copying, or multiple reproduction of any
material in this publication is permitted only under license from
the American Mathematical Society. Requests for such permission
should be addressed to the Assistant to the Publisher, American
Mathematical Society, P. O. Box 6248, Providence, Rhode Island
02940-6248. Requests can also be made by e-mail to
\textbf{\texttt{reprint-permission@ams.org.}}

\begin{center}
$\copyright$ 1999 by the American Mathematical Society. All rights reserved.\\
The American Mathematical Society retains all rights\\
except those granted to the United States Government.\\
Printed in the Unated States of America.
\end{center}

\begin{center}
$\circledinfty$\ The paper used in this book is acid-free and falls within the guidelines\\
established to ensure permanence and durability.\\
Visit the AMS home page at \textbf{URL:
\texttt{http://www.ams.org/}}
\end{center}

\begin{center}
10 9 8 7 6 5 4 3 2 \qquad 05 04 03 02 01 00
\end{center}
\end{copyrightpage}
