\begin{copyrightpage}
\begin{center}
\textbf{Editorial Board}
\end{center}

\begin{center}
David Bressoud \hfill Carl Pomerance\\
Robert Devaney, Chair \hfill Hung-Hsi Wu\\
\end{center}

\begin{center}
2000 \emph{Mathematics Subject Classification.} Primary 52C23;
Secondary 37B50, 37A25, 20H15, 82D25.
\end{center}

\begin{center}
The writing of this book was supported in part\\
by Texas grant ARP 003658-152 and NSF grant DMS 9531584.
\end{center}


\noindent \textsc{Abstract}. This is an interdisciplinary book aimed at senior level undergraduate mathematics students. The common thread throughout the book is a type of tiling, of which the best known example is the ``kite and dart" tiling of the plane. The book attempts to show that to understand this new type of structure it has been necessary to employ an unusual variety of specialties from mathematics and physics, and that this effort has resulted in new mathematics.\\

\hrulefill\\

\noindent \textbf{Library of Congress Cataloging-in-Publication
Data}

\noindent Radin, Charles, 1945--

Miles of tiles / Charles Radin.

\quad p. \quad cm. --- (Student mathematical library ; v. 1)

Includes bibliographical references and index.

ISBN 0-8218-1933-X (softcover : alk. paper)

1. Tiling (Mathematics)\quad I. Title.\quad II. Series.

\noindent QA166.8.R33 \quad 1999

\noindent 516--dc21 \hfill 99-20662

\hfill CIP\\

\hrulefill\\

\textbf{Copying and reprinting.} Individual readers of this
publication, and nonprofit libraries acting for them, are permitted
to make fair use of the material, such as to copy a chapter for use
in teaching or research. Permission is granted to quote brief
passages from this publication in reviews, provided the customary
acknowledgment of the source is given.

Republication, systematic copying, or multiple reproduction of any material in this publication is permitted only under license from the American Mathematical Society. Requests for such permission should be addressed to the Assistant to the Publisher, American Mathematical Society, P. O. Box 6248, Providence, Rhode Island 02940-6248. Requests can also be made by e-mail to \textbf{\texttt{reprint-permission@ams.org.}}\\

\begin{center}
$\copyright$ 1999 by the American Mathematical Society. All rights reserved.\\
The American Mathematical Society retains all rights\\
except those granted to the United States Government.\\
Printed in the United States of America.
\end{center}

\begin{center}
$\circledinfty$ The paper used in this book is acid-free and falls within the guidelines\\
established to ensure permanence and durability.\\
Visit the AMS home page at URL:
\textbf{\texttt{http://www.ams.org/}}
\end{center}

\begin{center}
10 9 8 7 6 5 4 3 2 \qquad 04 03 02 01 00
\end{center}
\end{copyrightpage}
